Exploratory Data Analysis with R by Martin Welss
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data from csv
wine <- read.csv("wineQualityWhites.csv")

# get some statistical info about the dataset
names(wine)
str(wine)
summary(wine)


```

# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots}

# what are the distributions?
# let's do some histograms

# quality is of of course the variable we want
# to predict from the others
ggplot(aes(x = quality), data = wine) + 
  geom_histogram()

# now the potential "input" variables
ggplot(aes(x = fixed.acidity), data = wine) + 
  geom_histogram()

ggplot(aes(x = residual.sugar), data = wine) + 
  geom_histogram()

ggplot(aes(x = total.sulfur.dioxide), data = wine) + 
  geom_histogram()

ggplot(aes(x = pH), data = wine) + 
  geom_histogram()

ggplot(aes(x = density), data = wine) + 
  geom_histogram()

ggplot(aes(x = sulphates), data = wine) + 
  geom_histogram()

ggplot(aes(x = alcohol), data = wine) + 
  geom_histogram()

# get better impression of distribution with frequnecy polygons
ggplot(aes(x = quality, y = ..count../sum(..count..)), data = wine) +
  geom_freqpoly() + 
  xlab('Quality') + 
  ylab('Percentage of observations with that quality')

```
```{r}

# alcohol seems the best lead, transform it to be a factor
wine$alcohol.bucket <- cut(wine$alcohol, c(8,9,10,11,12,13,14))

```
```{r echo=FALSE}

# and plot it as histogram
ggplot(aes(x = alcohol.bucket), data = wine) + 
  geom_histogram()

# and boxplot
qplot(x=alcohol.bucket, y=quality, data=wine, geom='boxplot')



```

# Univariate Analysis

### What is the structure of your dataset?
It's a tidy dataset where each variable is a column and each observation is a row. In this case, the dataset on white wines has 4898 observations of 13 variables.

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest in the dataset is quality and how to predict it from the other variables.  

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
From the plots we see that the quality resembles a normal distribution. So
we need to check the correlation between quality and the other variables.

### Did you create any new variables from existing variables in the dataset?
Yes, I created alcohol.bucket variable to get a feeling for the 
distributions of quality by alcohol and created also a corresponding boxplot. Quality seems to increase with more alcohol, but the correlation does not convince me, so this might be a dead end.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
The residual.sugar has a quite asymmetric distribution, the pH variable's distribution has two peaks. Other than adding the alcohol.bucket variable, 
I did no further adjustments. I needed the alcohol.bucket variable to create a boxplot of quality by alcohol. 



# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}

# let's create some scatterplots to find correlations...
ggplot(aes(x=quality, y=total.sulfur.dioxide), data=wine) + 
  geom_point(alpha=1/6)

ggplot(aes(x=quality, y=fixed.acidity), data=wine) + 
  geom_point(alpha=1/6)

ggplot(aes(x=quality, y=fixed.acidity*fixed.acidity), data=wine) + 
  geom_point(alpha=1/6)+ylim(c(0,120))

# no winner so far, maybe a linear model layer  helps

ggplot(aes(x=quality, y=alcohol), data=wine) + 
  geom_point(alpha=1/6) +
  geom_smooth(method='lm', color='red')

# is the correlation of higher order?
# try transformed variables

ggplot(aes(x=quality, y=sqrt(pH)), data=wine) + 
  geom_point(alpha=1/6)+
  geom_smooth(method='lm', color='red')

ggplot(aes(x=quality, y=fixed.acidity*total.sulfur.dioxide), data=wine) + 
  geom_point(alpha=1/10)+
  geom_smooth(method='lm', color='red')+
  ylim(c(0,2000))

```
```{r}

# create a new variable to combine acidity-types 
wine$prop.acidity <- wine$fixed.acidity / wine$volatile.acidity

```
```{r echo=FALSE}


# and plot it
ggplot(aes(x=quality, y=prop.acidity), data=wine) + 
  geom_point(alpha=1/8)+
  geom_smooth(method='lm', color='red')+
  ylim(c(0,75))

# what about correlations among the other variables?
ggplot(aes(x=alcohol, y=fixed.acidity), data=wine) +
  geom_point()
  
ggplot(aes(x=volatile.acidity, y=fixed.acidity), data=wine) +
  geom_point()


ggplot(aes(x=pH, y=alcohol), data=wine) +
  geom_point()


# we should at least here see a correlation
ggplot(aes(x=pH, y=fixed.acidity), data=wine) +
  geom_point() +
  geom_smooth(method='lm', color='red')


```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

Since we want to find the correlations between quality and the rest of the
variables, I created some scatterplots and tested the correlation between
promising quality-variable combinations. I had to deal with overplotting
by adding transparency to the points. Furthermore I added in some plots a geom_smooth layer based on a linear model to see the correlation more clearly.

Because I could not find any convincing relation, I tried transformations
of variables like the square fixed.acidity, the square root of pH and
introduced a new variable prop.acidity followed by corresponding plots.
But still no strong correlation could be found.

So I decided to test the some correlations numerically, but there is
none greater than 0.5, which would have been a breakthrough.

```{r}
cor.test(wine$quality, wine$alcohol)

cor.test(wine$quality, wine$total.sulfur.dioxide)

cor.test(wine$quality, wine$fixed.acidity)

cor.test(wine$quality, wine$prop.acidity)


```


### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

I found a slight linear relationship between pH and fixed.acidity where
the fixed acidity decreases with increasing pH values.

### What was the strongest relationship you found?

The strongest relationship so far is between quality and alcohol, but with a correlation coefficient of 0.44 the correlation is not strong enough for
a reliable prediction.


# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

# one more try with quality by alcohol
ggplot(aes(x=quality), data=wine)+
  geom_bar()+
  facet_wrap(~alcohol.bucket)

```
```{r}

# I need quality to be a factor, so create it
wine$quality.factor <- factor(wine$quality)

```
```{r echo=FALSE}


# use jitter to get a better color impression
ggplot(aes(x = pH, y = alcohol), data = wine) + 
  geom_jitter(aes(color = quality.factor))

# pH seems dead end, trying other variables
ggplot(aes(x = fixed.acidity, y = alcohol), data = wine) + 
  geom_jitter(aes(color = quality.factor))


ggplot(aes(x = total.sulfur.dioxide, y = alcohol), data = wine) + 
  geom_jitter(aes(color = quality.factor))


```
```{r}

# compute a new variable of promising candidates
wine$qmix <- wine$alcohol* 
  wine$total.sulfur.dioxide * 
  wine$fixed.acidity

```
```{r echo=FALSE}


# and create the plot
ggplot(aes(x=quality, y=qmix), data=wine) + 
  geom_point(alpha=1/8)+
  geom_smooth(method='lm', color='red')

```
```{r}

# and correlation test
cor.test(wine$quality, wine$qmix)


```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

To get a better impression of the relationship between quality and the
other variables, I needed quality to be a factor. So I created a new
factor variable from quality and used in the scatterplots to color
the points. Overplotting was again to be dealt with. But there was
no match winner revealed. I than created a new variable that was
the product of three other variables that so far showed at least
a little correlation with quality:

+ alcohol
+ total.sulfur.dioxide
+ fixed.acidity

The plot did not show a strong correlation, which was supported by 
the calculation of the correlation coefficient.

### Were there any interesting or surprising interactions between features?

What I found interesting was the relation between alcohol and quality
over several other variables like pH or fixed acidity.The relation
between alcohol and quality seems to be independent from the other variables since the higher quality points stay in the upper area of
the plot. This supports a constant correlation between alcohol and
quality.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}


ggplot(aes(x=quality), data=subset(wine, !is.na(alcohol.bucket)))+
  geom_bar(binwidth=0.5, fill="navy", colour="navy")+
  facet_wrap(~alcohol.bucket)+
  ggtitle("Quality distributions faceted by alcohol  [% by volume]")


```

### Description One
Sine quality is the output variable that we try to predict from the other 
variables, I present as first plot the quality distribution faceted by
alcohol, which so far has the strongest correlation with quality. We see
that the first two subplots with the low levels of alcohol have their maximum at quality 5 whereas  the following plots, the maximum moves to
higher quality levels.

### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(aes(x=quality, y=alcohol), data=wine) + 
  geom_point(alpha=1/6, color="navy") +
  geom_smooth(method='lm', color='red')+
  ggtitle("Scatterplot of quality by alcohol with linear model")+
  ylab("alcohol [% by volume]")


```

### Description Two
The second plot is a scatterplot which shows clearly a linear correlation between quality and alcohol with the red line depicting the correlation. This supports the assumption that higher levels of alcohol correspond with higher quality. The plot also shows that the vast majority of vines have mediocre quality and only very few have the top level quality of 9 or a bad quality below 4.

### Plot Three
```{r echo=FALSE, Plot_Three}

ggplot(aes(x = total.sulfur.dioxide, y = alcohol), data = wine) + 
  geom_jitter(aes(color = quality.factor), alpha=1/2) +
  xlim(c(0,300)) +
  ggtitle("Scatterplot of alcohol by sulfur dioxide colored by quality")+
  xlab("total sulfur dioxide [mg / dm^3]")+
  ylab("alcohol [% by volume]")

```

### Description Three

Plot three finally shows a scatterplot of alcohol by total sulfur dioxide colored by quality. I chose these three variables because they reflect
the key characteristics of my analysis:

1. alcohol and quality correlate
2. total sulfur dioxide and quality correlate 
3. the colors of quality in the plot support that

We see the lighter colors of better quality in the upper areas, especially
the purple points of high quality.

------

# Reflection

I started off with histograms of the variables to get an overview of the
distributions. Alcohol seemed most promising, so I wanted to create a 
boxplot. For that I needed alcohol to be a factor variable, so I used the
cut function to create the alcohol.bucket variable. The boxplot then
supported the relationship between alcohol and quality.

The bivariate analysis was somewhat sobering because no strong correlation
could be found.To avoid overseeing a correlation I added to some of the plots a linear model smoother but no correlation was stronger than between
alcohol and quality. Some slight correlation was found between sulfur dioxide, acidity and pH on the one hand and quality on the other hand.
I created a new variable and tried transformations and combinations of the
input variables.

The trend continued in the multivariate analysis.First I needed quality to be a factor, so I created a new variable for that. Then I used this factor
variable to color some of the scatter plots. I finally created another
variable the was the multiplication of alcohol, total.sulfur.dioxide and
fixed.acidity. But no strong correlation surfaced.

## Conclusion

I could not find a convincing correlation between quality and any
of the other variables. I am pretty sure after my analysis, that there
is at least no linear relationship. It could be a higher order 
relationship like exponential or polynomial, but I did not find any hints for that.
So from my analysis I would conclude that there must exist other variables
that correlate with quality of wine that were not in the dataset like the type of grape, geographical region, name of the vineyard or vintage. It would be most interesting to have those variables added to the dataset for
a future analysis.

If I were a winemaker, I would probably be more interested in the
relationship of the vintage and the quality _before_ I start the wine press.

