Exploratory Data Analysis with R by Martin Welss
========================================================

In this analysis I explore a dataset of white wine. Each observation contains a set
of objective, mostly chemical tests and quality variable which is based on sensory
data (median of at least 3 evaluations made by wine experts).

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
```

First, we load the dataset from the csv-file and 
get some statistical info about the dataset.


```{r echo=FALSE, Load_the_Data}
wine <- read.csv("wineQualityWhites.csv")

names(wine)
str(wine)
summary(wine)


```

# Univariate Plots Section

Quality is of course of primary interest because it 
is the variable we want to predict from the others,
so we look at its histogram first.

```{r echo=FALSE, Univariate_Plots}
ggplot(aes(x = quality), data = wine) + 
  geom_histogram()

```

We see that the quality is close to a normal distribution which is not fully symmetric because the value for quality 5 is higher than the corresponding value 7. Given that
the quality ranges from 1 to 10, we see the maximum at 6, which is sightly above
full mediocrity.

Now the histograms of the potential "input" variables:
```{r echo=FALSE}
ggplot(aes(x = fixed.acidity), data = wine) + 
  geom_histogram()
```

Nothing special about the histogram for fixed acidity: it resembles a normal distribution with no gaps and the values are in the range between 4 and 11. 

```{r echo=FALSE}

ggplot(aes(x = residual.sugar), data = wine) + 
  geom_histogram()+
  scale_x_log10()
```

The histogram on residual sugar has some interesting features:

+ I scaled the x-axis by log10 to have a tighter view
+ from left to right it has a peak, then a low and then two peaks
+ all three peaks are about the same height
+ it does not resemble any standard distribution I know

```{r echo=FALSE}

ggplot(aes(x = total.sulfur.dioxide), data = wine) + 
  geom_histogram()
```

The histogram for total sulfur dioxide looks again like a normal distribution, but other than fixed acidity the values have a much greater spectrum and range from 0 to over 300 with some outliers well over 400.

```{r echo=FALSE}

ggplot(aes(x = pH), data = wine) + 
  geom_histogram()
```

The value range for pH is pretty narrow with most of the values between 3 and 3.5, so all values are chemically well in the acidity range and not alkaline (which is above 7). The distribution resembles a normal distribution with an unusual double peak.

```{r echo=FALSE}

ggplot(aes(x = density), data = subset(wine, density<1.01)) + 
  geom_histogram(binwidth=0.001)
```

Interesting about the density plot is the narrow value range with most of them
in the interval between 0.9 and 1.0. The distribution looks like a slightly asymmetrical
normal distribution.

```{r echo=FALSE}

ggplot(aes(x = sulphates), data = wine) + 
  geom_histogram()
```

The sulphates distribution has a small tail on the right and a low left of the peak.

```{r echo=FALSE}

ggplot(aes(x = alcohol), data = wine) + 
  geom_histogram()

```

The alcohol distribution looks interesting because it is asymmetric and starts with a peak count at about 600 on the left and falls sharply down to 200 and then slowly descents to zero with little ups and downs.

I think we can get a better impression of the quality distribution 
with frequency polygons:


```{r echo=FALSE}

ggplot(aes(x = quality, y = ..count../sum(..count..)), data = wine) +
  geom_freqpoly() + 
  xlab('Quality') + 
  ylab('Percentage of observations with that quality')

```

Here the frequency polygons confirm the resemblance of a normal distribution.

The distributions of alcohol and quality resemble slightly 
so alcohol seems to be the best lead. We transform it to be a factor 
and remove the observations where alcohol.bucket is NA to get cleaner plots.
I want to get a more detailed impression of alcohol.bucket, so I plot a
histogram and a boxplot.


```{r}


wine$alcohol.bucket <- cut(wine$alcohol, c(8,9,10,11,12,13,14))

wine <- subset(wine, !is.na(wine$alcohol.bucket))


```
```{r echo=FALSE}

# plot it as histogram
ggplot(aes(x = alcohol.bucket), data = wine) + 
  geom_histogram()
```

The alcohol bucket histogram underlines the asymmetric outline of the distribution
with the peak already at interval (9,10] and the decreasing continuously to 
interval (13,14]

```{r echo=FALSE}

# and boxplot
qplot(x=alcohol.bucket, y=quality, data=wine, geom='boxplot')

```

The boxplot is the first highlight showing a clear relation between alcohol and quality with quality continuously increasing along with the level of alcohol.

# Univariate Analysis

### What is the structure of your dataset?
It's a tidy dataset where each variable is a column and each observation is a row. In this case, the dataset on white wines has 4898 observations of 13 variables.

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest in the dataset is quality and how to predict it from the other variables.  

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
From the plots we see that the quality resembles a normal distribution. So
we need to check the correlation between quality and the other variables. From
the alcohol-quality boxplot we can see a trend that Quality increases with more alcohol, but the correlation does not convince me yet, this needs more investigation in the
following sections.

### Did you create any new variables from existing variables in the dataset?
Yes, I created alcohol.bucket variable to get a feeling for the 
distributions of quality by alcohol and created also a corresponding boxplot. 

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
The residual.sugar has a quite asymmetric distribution, the pH variable's distribution has two peaks. Other than adding the alcohol.bucket variable, 
I needed the alcohol.bucket variable to create a boxplot of quality by alcohol and removed the records where alcohol.bucket was NA. 



# Bivariate Plots Section

To start, let's create some scatterplots to find interesting correlations...
 
```{r echo=FALSE, Bivariate_Plots}

ggplot(aes(x=quality, y=total.sulfur.dioxide), data=wine) + 
  geom_jitter(alpha=1/6)
```

I had to apply jitter and alpha to cope with overplotting, but there is no
clear relation between total sulfur dioxide and quality.

```{r echo=FALSE}
ggplot(aes(x=quality, y=fixed.acidity), data=wine) + 
  geom_jitter(alpha=1/6)
```

No clear trend either between fixed acidity and quality.

```{r echo=FALSE}

ggplot(aes(x=quality, y=fixed.acidity*fixed.acidity), data=wine) + 
  geom_jitter(alpha=1/6)+ylim(c(0,120))

```

The combination of the two acidity attributes just stretches the y-values over
a greater range compared to fixed acidity alone, but there is still no clear relation visible.

No winner so far, maybe a linear model layer helps

```{r}
ggplot(aes(x=quality, y=alcohol), data=wine) + 
  geom_jitter(alpha=1/6) +
  geom_smooth(method='lm', color='red')

```

Well, that looks promising let's keep the pair alcohol and quality in mind. We can
see clearly that quality increases with higher levels of alcohol.

Are there correlations of higher order?
Let's try transformed variables.


```{r echo=FALSE}

ggplot(aes(x=quality, y=sqrt(pH)), data=wine) + 
  geom_jitter(alpha=1/6)+
  geom_smooth(method='lm', color='red')
```

Nothing of interest between sqrt(pH) and quality, the line is almost flat.

```{r echo=FALSE}

ggplot(aes(x=quality, y=fixed.acidity*total.sulfur.dioxide), data=wine) + 
  geom_jitter(alpha=1/10)+
  geom_smooth(method='lm', color='red')+
  ylim(c(0,2000))

```

The smoother line for quality by the product of fixed acidity and total sulfur dioxide
shows a little slope, but it is far from a convincing correlation. We should 
nevertheless keep it in mind.

I create a new variable prop.acidity to combine acidity-types and plot it. 

```{r}

wine$prop.acidity <- wine$fixed.acidity / wine$volatile.acidity

```
```{r echo=FALSE}


ggplot(aes(x=quality, y=prop.acidity), data=wine) + 
  geom_jitter(alpha=1/8)+
  geom_smooth(method='lm', color='red')+
  ylim(c(0,75))

```

The correlation between quality and the combined acidities is visible as the slowly increasing line, but it is not strong enough nor is it better than fixed acidity alone.

What about correlations between the other variables?

```{r echo=FALSE}


ggplot(aes(x=alcohol, y=fixed.acidity), data=wine) +
  geom_jitter(alpha=1/8)
  
```

We see the values for fixed acidity stay in a range between 5 and 11 with more datapoints
in the area of lower alcohol. No correlation to see here. 

```{r echo=FALSE}
ggplot(aes(x=volatile.acidity, y=fixed.acidity), data=wine) +
  geom_jitter(alpha=1/8)


```

The scatterplot of the two acidities yields the shape of an egg in the left lower quadrant of the plot. 

```{r echo=FALSE}
ggplot(aes(x=pH, y=alcohol), data=wine) +
  geom_jitter(alpha=1/8)

```

The scatterplot of alcohol and pH results in quite equally distributed points which
shows no correlation at all.The plot is a little more dense for alcohol values below 10.

We should at least here see a correlation between pH and acidity because
of the chemical definition of pH.

```{r echo=FALSE}

ggplot(aes(x=pH, y=fixed.acidity), data=wine) +
  geom_jitter(alpha=1/8) +
  geom_smooth(method='lm', color='red')


```

We don't get disappointed as we see that fixed acidity decreases with higher pH values but that does not help for the prediction of quality.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

Since we want to find the correlations between quality and the rest of the
variables, I created some scatterplots and tested the correlation between
promising quality-variable combinations. I had to deal with overplotting
by adding transparency to the points. I carefully selected the alpha value so that the points of high quality observations are still visible. Furthermore I added in some plots a geom_smooth layer based on a linear model to see the correlation more clearly.

Because I could not find any convincing relation, I tried transformations
of variables like the square fixed.acidity, the square root of pH and
introduced a new variable prop.acidity followed by corresponding plots.
But still no strong correlation could be found.

So I decided to test the some correlations numerically, but there is
none greater than 0.5, which would have been a breakthrough.

```{r}
cor.test(wine$quality, wine$alcohol)

cor.test(wine$quality, wine$total.sulfur.dioxide)

cor.test(wine$quality, wine$fixed.acidity)

cor.test(wine$quality, wine$prop.acidity)


```


### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

I found a slight linear relationship between pH and fixed.acidity where
the fixed acidity decreases with increasing pH values.

### What was the strongest relationship you found?

The strongest relationship so far is between quality and alcohol, but with a correlation coefficient of 0.44 the correlation is not strong enough for
a reliable prediction.


# Multivariate Plots Section

We start with the strongest lead alcohol and quality but this time 
faceted by the different levels of alcohol using the alcohol.bucket variable. 

```{r echo=FALSE, Multivariate_Plots}

# one more try with quality by alcohol
ggplot(aes(x=quality), data=wine)+
  geom_bar()+
  facet_wrap(~alcohol.bucket)

```

The subplots show clearly how different the counts are regarding the alcohol levels:
there are only a few (less than 100) datapoints for the highest alcohol level of 13 whereas for the levels of 9 and 10 there are more than 1500 datapoints. But plot
also confirms the correlation between alcohol and quality as we see the bars wander
to the right towards better quality with higher alcohol levels.

I think I need quality to be a factor, so create it

```{r}

wine$quality.factor <- factor(wine$quality)

```

I use jitter to get a better color impression.

```{r echo=FALSE}


ggplot(aes(x = pH, y = alcohol), data = wine) + 
  geom_point(aes(color = quality.factor))+
  scale_color_brewer(type = "seq")

```

We see here again that the darker points of better quality correspond with higher
alcohol levels, but pH is equally distributed among quality.
pH seems to be a dead end, let's try other variables.

```{r echo=FALSE}
ggplot(aes(x = fixed.acidity, y = alcohol), data = wine) + 
  geom_point(aes(color = quality.factor))+
  scale_color_brewer(type = "seq")
```

Same conclusion with fixed acidity: the alcohol-quality relation shows again,
but fixed acidity has no influence.

```{r echo=FALSE}

ggplot(aes(x = total.sulfur.dioxide, y = alcohol), data = wine) + 
  geom_point(aes(color = quality.factor))+
  scale_color_brewer(type = "seq")


```

The plot with total sulfur dioxide finally shows a very small interesting
deformation where the lighter of low quality datapoints are shifted to the lower right.
We should keep that in mind.

We compute a new variable of promising candidates and create a corresponding plot.

```{r}

wine$qmix <- wine$alcohol* 
  wine$total.sulfur.dioxide * 
  wine$fixed.acidity

```

The new qmix variable is quite uniformly distributed around the center of the plot,
so no new revelations here.

```{r echo=FALSE}


ggplot(aes(x=quality, y=qmix), data=wine) + 
  geom_jitter(alpha=1/8)+
  geom_smooth(method='lm', color='red')

```

Finally we test the correlation, but still not better than alcohol. 

```{r}

cor.test(wine$quality, wine$qmix)

```


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

To get a better impression of the relationship between quality and the
other variables, I needed quality to be a factor. So I created a new
factor variable from quality and used in the scatterplots to color
the points. Overplotting was again to be dealt with and again I selected the alpha value so that the colors of the high and low quality observations stay visible. But there was
no match winner revealed. I than created a new variable that was
the product of three other variables that so far showed at least
a little correlation with quality:

+ alcohol
+ total.sulfur.dioxide
+ fixed.acidity

The plot did not show a strong correlation, which was supported by 
the calculation of the correlation coefficient.

### Were there any interesting or surprising interactions between features?

What I found interesting was the relation between alcohol and quality
over several other variables like pH or fixed acidity.The relation
between alcohol and quality seems to be independent from the other variables since the higher quality points stay in the upper area of
the plot. This supports a constant correlation between alcohol and
quality.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}


ggplot(aes(x=quality), data=subset(wine, !is.na(alcohol.bucket)))+
  geom_bar(binwidth=0.5, fill="navy", colour="navy")+
  facet_wrap(~alcohol.bucket)+
  ggtitle("Quality distributions faceted by alcohol  [% by volume]")


```

### Description One
Sine quality is the output variable that we try to predict from the other 
variables, I present as first plot the quality distribution faceted by
alcohol, which so far has the strongest correlation with quality. We see
that the first two subplots with the low levels of alcohol have their maximum at quality 5 whereas  the following plots, the maximum moves to
higher quality levels. I think this plot is important and quite interesting, because it gives a good impression of the relation between the quality and alcohol which is its strongest correlation. Very few wines have excellent quality and so have only a few wines a bad quality.

### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(aes(x=quality, y=alcohol), data=wine) + 
  geom_jitter(alpha=1/6, color="navy") +
  geom_smooth(method='lm', color='red')+
  ggtitle("Scatterplot of quality by alcohol with linear model")+
  ylab("alcohol [% by volume]")


```

### Description Two
The second plot is a scatterplot which shows clearly a linear correlation between quality and alcohol with the red line depicting the correlation. This supports the assumption that higher levels of alcohol correspond with higher quality. The plot also shows that the vast majority of vines have mediocre quality and only very few have the top level quality of 9 or a bad quality below 4.

### Plot Three
```{r echo=FALSE, Plot_Three}

ggplot(aes(x = total.sulfur.dioxide, y = alcohol), data = wine) + 
  geom_jitter(alpha=1/8, color="navy") +
  scale_color_brewer(type = "seq") +
  facet_wrap(~quality)+
  xlim(c(0,300)) +
  ggtitle("Scatterplot of alcohol by sulfur dioxide colored by quality")+
  xlab("total sulfur dioxide [mg / dm^3]")+
  ylab("alcohol [% by volume]")

```

### Description Three

Plot three finally shows a scatterplot of alcohol by total sulfur dioxide faceted by quality. I chose these three variables because they reflect
the key characteristics of my analysis:

1. alcohol and quality correlate
2. total sulfur dioxide and quality correlate 
3. the subplots by quality support that

We see over the subplots from 3 (poor quality) to 9 (best quality) how the
datapoints climb to higher alcohol levels between 10 and 12 % and a sulfur level 
condenses between 50 and 150 mg/dm³. We see also that most wines have a medium quality of 5 or 6.


------

# Reflection

I started off with histograms of the variables to get an overview of the
distributions. Alcohol seemed most promising, so I wanted to create a 
boxplot. For that I needed alcohol to be a factor variable, so I used the
cut function to create the alcohol.bucket variable. The boxplot then
supported the relationship between alcohol and quality.

The bivariate analysis was somewhat sobering because no strong correlation
could be found.To avoid overseeing a correlation I added to some of the plots a linear model smoother but no correlation was stronger than between
alcohol and quality. Some slight correlation was found between sulfur dioxide, acidity and pH on the one hand and quality on the other hand.
I created a new variable and tried transformations and combinations of the
input variables.

The trend continued in the multivariate analysis.First I needed quality to be a factor, so I created a new variable for that. Then I used this factor
variable to color some of the scatter plots. I finally created another
variable the was the multiplication of alcohol, total.sulfur.dioxide and
fixed.acidity. But no strong correlation surfaced.

## Conclusion

I could not find a convincing correlation between quality and any
of the other variables. I am pretty sure after my analysis, that there
is at least no linear relationship. It could be a higher order 
relationship like exponential or polynomial, but I did not find any hints for that.
So from my analysis I would conclude that there must exist other variables
that correlate with quality of wine that were not in the dataset like the type of grape, geographical region, name of the vineyard or vintage. It would be most interesting to have those variables added to the dataset for
a future analysis.

If I were a winemaker, I would probably be more interested in the
relationship of the vintage and the quality _before_ I start the wine press.

